---
title: "Using the tidyFIA package"
author: "Brian Clough & Henry Rodman"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using the tidyFIA package}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  comment = NA,
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  dev = "png",
  dpi = 150,
  fig.asp = 0.8,
  fig.width = 5,
  out.width = "90%",
  fig.align = "center"
)
```

This vignette demonstrates the basic use of the tidyFIA package. The core function is `tidy_fia()`, which queries the FIA database by a single state abbreviation, a string of state abbreviations, or a polygonal area of interest supplied as an `sf` object. In this example, we will use {tidyFIA} to extract recent FIA data for the Adirondack & Green Mountains ecological supersection, which spans the states of New York and Vermont in the northeastern US.

```{r libs}
library(dplyr)
library(ggplot2)
library(sf)
library(tidyr)

library(tidyFIA)

ggplot2::theme_set(theme_bw())
```

First we will download the supersection shapefile and create an `aoi` to pass to tidyFIA.

```{r supersection, include=TRUE}

# Download master shapefile
super_zip <- file.path(
  tempdir(),
  "gis-supersection-shape-file.zip"
)

super_url <- "https://www.arb.ca.gov/cc/capandtrade/protocols/usforest/2014/supersectionshapefiles/gis-supersection-shape-file.zip"

download.file(
  url = super_url,
  destfile = super_zip
)

unzip(
  zipfile = super_zip,
  exdir = tempdir()
)

super_file <- file.path(tempdir(), "Supersections/Supersections.shp")

# import and reproject to EPSG:4326
supersectionShape <- st_read(dsn = super_file) |>
  st_transform(crs = 4326) |>
  select(SSection, geometry)

# filter
ecoreg <- supersectionShape |>
  filter(SSection == "Adirondacks & Green Mountains")
```

The function `tidy_fia` can either query a PostGIS database hosted by SilviaTerra (`postgis` = TRUE) or download the relevant FIA csv tables from [FIADB](https://apps.fs.usda.gov/fia/datamart/CSV/datamart_csv.html). Querying the database is generally faster than downloading and importing the csv files. The plots are filtered down to the points within the spatial area of interest (`aoi`). The `table_names` argument can be used to specify which FIADB tables from the "Index of Tables" in the [FIADB User Guide](https://www.fia.fs.fed.us/library/database-documentation/current/ver80/FIADB%20User%20Guide%20P2_8-0.pdf) you would like to download. The imported tables are returned as a list of dataframes.

```{r tidy_fia}
fia_data <- tidy_fia(
  aoi = ecoreg,
  postgis = TRUE,
  table_names = c("plot", "tree", "subplot", "cond")
)
```

A simple map of the plot locations can be printed by calling `plot(fia_data)`. The `plot` method for the `tidyFIA` class will generate a {ggplot2} map that can be added to other plot elements such as some nearby state boundaries to add geographic context.

```{r map, include = TRUE}
p <- plot(fia_data)

nearby_states <- spData::us_states |>
  st_transform(st_crs(ecoreg)) |>
  st_join(ecoreg) |>
  filter(!is.na(SSection))

p +
  geom_sf(
    data = nearby_states,
    color = "blue",
    alpha = 0
  )
```

### Tree data
The 'tree' table will be stored as an element in the list: `fia_data[["tree"]]`. This table can be used to explore tree-level attributes such as height:diameter relationships among species.
```{r trees, include = TRUE}
# get REF_SPECIES tables
ref_species <- read_ref_table("REF_SPECIES") |>
  select(spcd, common_name)

# most common species
common_spp <- fia_data[["tree"]] |>
  group_by(spcd) |>
  tally() |>
  top_n(9, wt = n) |>
  pull(spcd)

fia_data[["tree"]] |>
  filter(spcd %in% common_spp) |>
  left_join(ref_species) |>
  ggplot(aes(x = dia, y = ht)) +
  geom_point(alpha = 0.2) +
  facet_wrap(~common_name)
```

### Plot attributes
The tree-level data in the 'tree' table can be summarized to obtain plot-level summaries.
```{r plot_attributes, include = TRUE}
plot_stats <- fia_data[["tree"]] |>
  group_by(plt_cn) |>
  summarize(
    bapa = sum(tpa_unadj * 0.005454 * dia^2, na.rm = TRUE),
    tpa = sum(tpa_unadj, na.rm = TRUE),
    qmd = sqrt(bapa / tpa / 0.005454)
  ) |>
  full_join(
    fia_data[["plot"]] |> select(cn, invyr),
    by = c("plt_cn" = "cn")
  ) |>
  replace_na(replace = list(bapa = 0, tpa = 0, qmd = 0))
```

The plot-level attributes could be displayed using histograms. In this example, we are looking at the distribution of basal area per acre, trees per acre, and quadratic mean diameter in plots measured in 2014 or later.

```{r histograms, include = TRUE}
plot_stats |>
  filter(invyr >= 2014) |>
  pivot_longer(
    cols = c("bapa", "tpa", "qmd"),
    names_to = "attribute",
    values_to = "value"
  ) |>
  ggplot(aes(x = value)) +
  geom_histogram() +
  facet_wrap(~attribute, scales = "free")
```

Distributions of plot-level observations can be plotted with respect to time (`invyr`).

```{r time}
plot_stats |>
  pivot_longer(
    cols = c("bapa", "tpa", "qmd"),
    names_to = "attribute",
    values_to = "value"
  ) |>
  ggplot(aes(x = invyr, y = value)) +
  geom_point(alpha = 0.2) +
  geom_smooth() +
  facet_wrap(~attribute, scales = "free")
```

You may also be interested in plotting observations spatially.
```{r qmd_map}
plot_stats_spatial <- plot_stats |>
  rename(geometry = geom) |>
  st_as_sf()

ggplot() +
  geom_sf(
    data = nearby_states,
    color = "blue",
    alpha = 0
  ) +
  geom_sf(
    data = ecoreg,
    color = "black",
    alpha = 0
  ) +
  geom_sf(
    data = plot_stats_spatial |> arrange(qmd),
    aes(color = qmd, geometry = geometry)
  ) +
  scale_color_gradientn(colors = terrain.colors(10))
```
