---
title: "White Spruce in Northern MN"
author: "Henry Rodman"
vignette: >
  %\VignetteIndexEntry{White spruce}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  comment = NA,
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  dev = "png",
  dpi = 150,
  fig.asp = 0.8,
  fig.width = 6,
  out.width = "90%",
  fig.align = "center"
)
```

```{r libs}
library(dplyr)
library(ggplot2)
library(sf)
library(tidyr)

library(tidyFIA)

ggplot2::theme_set(theme_bw())
```

White spruce is a commercially and ecologically important species in northern Minnesota so I am interested in understanding more about the way it tends to grow across the region.

FIA data is useful for this type of quest so I am going to use `tidy_fia` to download and import the data so I can spend my time on the part of the process that I care about: learning about white spruce!

## Spatial data
There is a shapefile for just about everything on the internet, including one for all counties in Minnesota.
I googled "minnesota counties shapefile" and found the one I want on one of the first links (hint: if you see a download link button you can right-click and copy the web address where the file is stored).
I can use R to download and read the zipped shapefile, then filter the file down to the counties where I want to know more about white spruce
```{r counties}
counties_url <- "ftp://ftp.gisdata.mn.gov/pub/gdrs/data/pub/us_mn_state_dnr/bdry_counties_in_minnesota/shp_bdry_counties_in_minnesota.zip"

counties_zip <- file.path(
  tempdir(),
  "shp_bdry_counties_in_minnesota.zip"
)

download.file(
  url = counties_url,
  destfile = counties_zip
)

unzip(
  zipfile = counties_zip,
  exdir = tempdir()
)

county_shp <- file.path(tempdir(), "mn_county_boundaries.shp")

counties <- st_read(county_shp) %>%
  st_transform(4326)
  
county_names <- c(
  "Cook",
  "Lake",
  "St. Louis",
  "Koochiching",
  "Itasca"
)

northern_counties <- counties %>%
  filter(CTY_NAME %in% county_names) %>%
  group_by(CTY_NAME, CTY_FIPS) %>%
  summarize() %>%
  ungroup()
```

Here is a map for some context:
```{r}
mn <- spData::us_states %>%
  filter(NAME == "Minnesota")
  
ggplot() +
  geom_sf(
    data = mn,
    color = "black",
    alpha = 0
  ) +
  geom_sf(
    data = northern_counties,
    color = "black",
    fill = "darkgreen"
  ) +
  labs(
    title = "Northern MN"
  )
```

## FIA data

Now I will collect the FIA "plot" and "tree" tables for Minnesota using the function `tidy_fia`.

```{r get_data}
aoi <- northern_counties %>%
  summarize()

fia_data <- tidy_fia(
  aoi = aoi,
  table_names = c("plot", "tree"),
  postgis = TRUE
)
```

Since the tree table does not have common names for the trees I am importing the REF_SPECIES table from FIADB using the function `read_ref_table`.
```{r ref_species}
ref_species <- read_ref_table("REF_SPECIES") %>%
  select(spcd, common_name, sftwd_hrdwd)

```

The tree table has codes for state and county but the readable names are stored in a separate table ("COUNTY") which is stored at the state-level but is also available for the entire US.
```{r}
ref_county <- read_ref_table("COUNTY") %>%
  select(statecd, countycd, county = countynm)
```

## Analysis

Perhaps white spruce shows differences in height and diameter relationships with respect to the type of forest community it inhabits.
I am going to designate each plot as 'hardwood', 'mixed', or 'softwood' using plot basal area in hardwood/softwood trees in the tree table.

```{r plot_attributes}
plot_stats <- fia_data[["tree"]] %>%
  left_join(ref_species) %>%
  group_by(plt_cn, sftwd_hrdwd) %>%
  summarize(
    bapa = sum(tpa_unadj * 0.005454 * dia ^ 2, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  pivot_wider(
    names_from = "sftwd_hrdwd",
    values_from = "bapa"
  ) %>%
  replace_na(
    replace = list(H = 0, S = 0)
  ) %>%
  mutate(
    total = H + S,
    type = case_when(
      H / (H + S) > 0.7 ~ "hardwood",
      H / (H + S) >= 0.3 ~ "mixed",
      H / (H + S) < 0.3 ~ "softwood"
    )
  )

```

Now I can filter the TREE table down to just live (`statuscd == 1`) white spruce.
```{r tree_ht}
white_spruce <- fia_data[["tree"]] %>%
  left_join(ref_species, by = "spcd") %>%
  filter(
    common_name == "white spruce",
    statuscd == 1, !is.na(dia), !is.na(ht)
  ) %>%
  left_join(plot_stats, by = "plt_cn") %>%
  left_join(ref_county, by = c("statecd", "countycd"))
```

To get a sense for height/diameter relationships in white spruce I will make a graph that shows the trend by forest type and county.
```{r county_type_plot}
white_spruce %>%
  ggplot(aes(x = dia, y = ht, color = type)) +
  facet_wrap(~ county) +
  geom_smooth() +
  labs(
    x = "DBH (in)",
    y = "total height (ft)",
    title = "white spruce in northern MN counties",
    caption = "smoothed relationship between DBH and total height across all\nwhite spruce observed on FIA plots since 2000"
  )

```

That graph is somewhat hard to read and does not show any obvious differences in height:diameter relationships so I will try one graph with a separate line for each county.
```{r county_plot}
white_spruce %>%
  ggplot(aes(x = dia, y = ht, color = county)) +
  geom_smooth() +
  labs(
    x = "DBH (in)",
    y = "total height (ft)",
    title = "white spruce in northern MN counties",
    caption = "smoothed relationship between DBH and total height\nacross all white spruce observed on FIA plots since 2000"
  )

```
