---
title: "White Spruce in Northern MN"
author: "Henry Rodman"
vignette: >
  %\VignetteIndexEntry{White spruce}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  comment = NA,
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  dev = "png",
  dpi = 150,
  fig.asp = 0.8,
  fig.width = 5,
  out.width = "90%",
  fig.align = "center"
)
```

```{r libs}
library(dplyr)
library(ggplot2)
library(sf)
library(tidyr)

library(tidyFIA)

ggplot2::theme_set(theme_bw())
```

White spruce is a commercially and ecologically important species in northern Minnesota so I am interested in understanding more about the way it tends to grow across the region.

FIA data is useful for this type of quest so I am going to use `tidy_fia` to download and import the data so I can spend my time on the part of the process that I care about: learning about white spruce!

## Spatial data
There is a shapefile for just about everything on the internet, including one for all counties in Minnesota.
I googled "minnesota counties shapefile" and found the one I want on one of the first links (hint: if you see a download link button you can right-click and copy the web address where the file is stored).
I can use R to download and read the zipped shapefile, then filter the file down to the counties where I want to know more about white spruce
```{r counties}
counties_url <- "ftp://ftp.gisdata.mn.gov/pub/gdrs/data/pub/us_mn_state_dnr/bdry_counties_in_minnesota/shp_bdry_counties_in_minnesota.zip"

counties_zip <- file.path(
  tempdir(),
  "shp_bdry_counties_in_minnesota.zip"
)

download.file(
  url = counties_url,
  destfile = counties_zip
)

unzip(
  zipfile = counties_zip,
  exdir = tempdir()
)

county_shp <- file.path(tempdir(), "mn_county_boundaries.shp")

counties <- st_read(county_shp) %>%
  st_transform(4326)
  
county_names <- c(
  "Cook",
  "Lake",
  "St. Louis",
  "Koochiching",
  "Itasca",
  "Carlton",
  "Aitkin",
  "Cass",
  "Crow Wing",
  "Hubbard",
  "Beltrami",
  "Lake of the Woods"
)

northern_counties <- counties %>%
  filter(CTY_NAME %in% county_names) %>%
  group_by(CTY_NAME, CTY_FIPS) %>%
  summarize()
```

Here is a map for some context:
```{r}
mn <- spData::us_states %>%
  filter(NAME == "Minnesota")
  
ggplot() +
  geom_sf(
    data = mn,
    color = "black",
    alpha = 0
  ) +
  geom_sf(
    data = northern_counties,
    color = "black",
    fill = "darkgreen"
  ) +
  labs(
    title = "Northern MN"
  )
```

## FIA data

Now I will download the FIA "PLOT" and "TREE" tables for Minnesota using the function `tidy_fia`.
```{r get_data}
aoi <- northern_counties %>%
  summarize()

fia_data <- tidy_fia(
  state = "MN",
  files = c("PLOT", "TREE")
)
```

Since the TREE table does not have common names for the trees I am importing the REF_SPECIES table from FIADB using the function `read_ref_table`s
```{r ref_species}
ref_species <- read_ref_table("REF_SPECIES") %>%
  select(SPCD, COMMON_NAME, SFTWD_HRDWD)

```

## Analysis

We are only interested in white spruce in northern MN so I am filtering the FIA plots down to ones measured in the northern counties (as identified by the county FIPS codes in the PLOT table).
```{r filter_plots}
plots <- fia_data[["PLOT"]] %>%
  filter(
    COUNTYCD %in% northern_counties$CTY_FIPS,
    INVYR >= 2000
  ) %>%
  transmute(
    CN,
    COUNTYCD,
    CTY_NAME = northern_counties$CTY_NAME[match(COUNTYCD, northern_counties$CTY_FIPS)]
  )
```

Perhaps white spruce shows differences in height and diameter relationships with respect to the type of forest community it inhabits.
I am going to designate each plot as 'hardwood', 'mixed', or 'softwood' using plot basal area in hardwood/softwood trees in the TREE table.
```{r plot_attributes}

plot_stats <- fia_data[["TREE"]] %>%
  filter(PLT_CN %in% plots$CN) %>%
  left_join(ref_species, by = "SPCD") %>%
  group_by(PLT_CN, SFTWD_HRDWD) %>%
  summarize(
    bapa = sum(TPA_UNADJ * 0.005454 * DIA ^ 2, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  pivot_wider(
    names_from = "SFTWD_HRDWD",
    values_from = bapa
  ) %>%
  replace_na(
    replace = list(H = 0, S = 0)
  ) %>%
  mutate(
    total = H + S,
    type = case_when(
      H / (H + S) > 0.7 ~ "hardwood",
      H / (H + S) >= 0.3 ~ "mixed",
      H / (H + S) < 0.3 ~ "softwood"
    )
  ) %>%
  left_join(
    plots,
    by = c("PLT_CN" = "CN")
  )

```

Now I can filter the TREE table down to just live (`STATUSCD == 1`) white spruce.
```{r tree_ht}
white_spruce <- fia_data[["TREE"]] %>%
  left_join(ref_species, by = "SPCD") %>%
  filter(
    PLT_CN %in% plot_stats$PLT_CN,
    COMMON_NAME == "white spruce",
    STATUSCD == 1, !is.na(DIA), !is.na(HT)
  ) %>%
  left_join(plot_stats, by = "PLT_CN"
  ) %>%
  select(
    PLT_CN, CTY_NAME, type, H, S, total,
    COMMON_NAME, DIA, HT
  )
```

To get a sense for height/diameter relationships in white spruce I will make a graph that shows the trend by forest type and county.
```{r county_type_plot}
white_spruce %>%
  ggplot(aes(x = DIA, y = HT, color = type)) +
  facet_wrap(~ CTY_NAME) +
  geom_smooth() +
  labs(
    x = "DBH (in)",
    y = "total height (ft)",
    title = "white spruce in northern MN counties",
    caption = "smoothed relationship between DBH and total height across all\nwhite spruce observed on FIA plots since 2000"
  )

```

That graph is somewhat hard to read and does not show any obvious differences in height:diameter relationships so I will try one graph with a separate line for each county.
```{r county_plot}
white_spruce %>%
  ggplot(aes(x = DIA, y = HT, color = CTY_NAME)) +
  geom_smooth() +
  labs(
    x = "DBH (in)",
    y = "total height (ft)",
    title = "white spruce in northern MN counties",
    caption = "smoothed relationship between DBH and total height\nacross all white spruce observed on FIA plots since 2000"
  )

```
